``Interestingness'' in Mathematical Theory Exploration

### 2018-05-17 PhDThesis/futurework.tex
Drag a bunch of stuff into thesis folder

\begin{sidewaysfigure}
  \centering
  \includegraphics[trim=3cm 2cm 3cm 2cm, width=1.00\textwidth]{Gantt.pdf}
  \caption{Gantt chart showing research direction, sub-tasks and approximate timeframes. CICM is the \emph{Conference on Intelligent Computer Mathematics}.}
  \label{fig:gantt}
\end{sidewaysfigure}

Our use of clustering to pre-process \qspec{} signatures has required many decisions and tradeoffs to be made. Hence our approach is just one possibility out of many alternatives which could be investigated to push this work further. In addition, there are other ways in which machine learning could aid theory exploration besides our relevance filter technique. The Gantt chart in Figure \ref{fig:gantt} shows how these relate to the short- and long-term direction being taken by this research. Below, we elaborate on the details, background and motivation for these choices.

@@

Approaches vary, from those which are directly related to theory exploration (such as the scheme-instantiating approach of \citep{Montano-Rivas.McCasland.Dixon.ea:2012}, which forms part of a theory exploration system), to others which are more closely related to theories of human learning and discovery \citep{Piantadosi.Tenenbaum.Goodman:2012, mullerunderstanding}. Those based on finding patterns in data, such as \citep{Wille:2005}, may be useful in tandem with our expression database, and the results of value generators like \qcheck{}.

Since tools like \hspec{} already call out to third-party automated theorem provers, and indeed our own ML4HS system uses the external Weka program, there may also be merit in using external concept formation or conjecture generation tools (or reimplementations of their ideas), in order to build up more structure on top of that provided by the code we are exploring. For example, the approaches taken by AM \citep{lenat1977automated, lenat1979automated}, Graffiti \citep{delavina2005some, delavina2005graffiti} and HR \citep{colton1999automatic, colton2000agent} could be used alongside those of \qspec{}.

@@

  \caption{The Wundt curve, reproduced from \citep{berlyne1970novelty}. The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}

@@

\emph{Artificial curiosity} (AC) describes active learning systems which are rewarded based on how interesting the input or data they discover is \citep{schmidhuber2006developmental}. Although framed in the context of \emph{reinforcement learning}, this reliance on interest is clearly relevant to our theory exploration setting.

As an unsupervised learning task, AC has no access to labels or meanings associated with its input; the only features it can learn are the structure and relationships inherent in the data, in a similar way to our recurrent clustering algorithm. The unifying principle of AC methods is to force systems away from inputs which are not amenable to learning; either because they are so familiar that there is nothing left to learn, or so unfamiliar that they are unintelligible. The resulting behaviour is characterised by the \emph{Wundt curve} (shown in Figure \ref{fig:wundt}) \footnote{In practice, many measures avoid negative values for simplicity, in which cases we replace all negative points on the curve with zero.}, which has been used in psychology to explain human aesthetics and preferences \citep{berlyne1970novelty}. This same behaviour may be applicable to the theorems produced by a theory exploration system.

@@

A framework encompassing many examples of the explicit approach is given in \citep{oudeyer2007intrinsic} in the context of reinforcement learning; for comparison, many similar measures are surveyed in a data mining context in \citep{geng2006interestingness}. Many more reinforcement learning examples can be found in \citep{Kaplan2006, Lipson2007, Luciw2011, Macedo2000, Ramik.Sabourin.Madani:2013, Roa.Kruijff.Jacobsson:2009, Schmidhuber:1991, oudeyer2004intelligent}; whilst more general descriptions are given in \citep{Schaul.Sun.Wierstra.ea:2011, Scott1989, maher2008achieving}, which may be more amenable to our theory exploration setting.

@@

Another example of explicit artificial curiosity is given in \citep{Hester.Stone:2012}, where world states which cause \emph{disagreement} among a population of decision trees (a \emph{random forest} \citep{randomforests}) are considered interesting. Since the models make stochastic predictions, the disagreement follows a Wundt curve as the complexity of state transitions increases: for parts of the state space which have been fully learned, the models will agree on accurate predictions; for parts which are unlearnable, the models cannot infer any structure, and will converge to reporting the same average value. Whilst the latter predictions may not be \emph{accurate}, they will be \emph{in agreement}, hence pushing down the interestingness of states which are too complex.

Many examples of the implicit case are based on \emph{coevolution}: rewarding one part of the system for exploiting another part, and vice versa. In \citep{Schmidhuber1999} a pair of learning algorithms place virtual ``bets'' on the outcome of actions, and the winner is rewarded at the expense of the loser. Due to the risk involved, each algorithm will only bet when it is confident in its prediction, and bets will only be actioned when each algorithm is confident in a \emph{different} outcome. The overall behaviour of this system is therefore similar to the explicit measure of disagreement used in the random forest example. In terms of theory exploration, such a scheme could be used to find theorems which are \emph{non-obvious}, and hence informative in some way to the user.

Another case is the ``darwinian brain'' of Fernando et al. \citep{fernando2013design1, fernando2013design2}. This coevolves a population of problem generators and problem solvers, rewarding the solvers based on their speed, and rewarding the generators based on the \emph{variance} of the solvers' speed. This avoids trivial problems (which all solvers can quickly overcome) and complex problems (which no solver can manage), and focuses on those with the most possibility for learning. It is easy to imagine such a general architecture being populated by conjecture generators and theorem provers to form a theory exploration system.

@@

Coevolution is a form of \emph{evolutionary computation}; an umbrella term for heuristic search algorithms which mimic the process of evolution by natural selection among a population of candidate solutions \citep{back1997evolutionary}. Whilst \emph{genetic algorithms} are perhaps the most well-known instance of evolutionary computation, their use of \emph{strings} to represent solutions causes complications when comparing to a domain like theory exploration, where recursive structures of unbounded depth arise. Thankfully these problems are not insurmountable, for example \emph{genetic programming} can operate on tree-structures natively \citep{banzhaf1998genetic}, which makes evolutionary computation a useful source of ideas for reuse in our theory exploration setting (there are also precedents for using evolutionary computation in a theorem proving domain \citep{spector2008genetic}).

@@

Pure exploration (i.e. for its own sake) has been studied in evolutionary computation for two main reasons: \emph{artificial life} and \emph{deceptive problems}. The former attempts to gain insight into the nature of life and biology through competition over limited resources. Whilst this may have utility in resource allocation, e.g. efficient scheduling of a portfolio of ATP programs, there is no direct connection to interestingness in theory exploration, so we will not consider it further (note that similar resource-usage ideas can also be found in the literature on \emph{artificial economies}, e.g. \citep{baum2000evolution}).

On the other hand, work on deceptive problems is highly relevant, as it has lead to studying various notions of intrinsic fitness, which are analogous to the interestingness measures we want. Deceptive problems are those where \textquote{pursuing the objective may prevent the objective from being reached} \citep{lehman2011abandoning}, which is caused by the fitness (objective) function having many local optima which are easy to find (e.g. by hill climbing), but few global optima which are hard to find. Many approaches try to avoid deception by augmenting the given fitness function to promote \emph{diversity} and \emph{novelty}, such as \emph{niching methods} \citep{sareni1998fitness}.

@@

Traditional selection methods are biased towards high fitness individuals (this is especially clear for truncation). Alternative schemes have been proposed which favour diversity \emph{at the expense of} fitness. For example, the fitness uniform selection scheme (FUSS) \citep{hutter2002fitness} selects a target fitness $f_t$ uniformly from the interval $\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the population. An individual $s$ is then selected with fitness closest to $f_t$, i.e. $s = \argmin\limits_{x} \lvert f(x) - f_t \rvert$

@@

Once we begin this process of augmenting fitness functions, or abandoning their use as objectives, an obvious question arises: what happens if our new function contains nothing of the original? This kind of pure exploration scenario leads to a variety of ideas for \emph{instrinsic} fitness, such as novelty \citep{lehman2011abandoning}, which can lead to learning useful ``stepping stones'' even in objective-driven domains. Such intrinsic notions of fitness are direct analogues of the interestingness measures we seek for theory exploration.

>>>>

Our use of clustering to pre-process \qspec{} signatures has required many
decisions and tradeoffs to be made. Hence our approach is just one possibility
out of many alternatives which could be investigated to push this work
further. In addition, there are other ways in which machine learning could aid
theory exploration besides our relevance filter technique. Below, we elaborate
on the details, background and motivation for these choices.

@@

Approaches vary, from those which are directly related to theory exploration
(such as the scheme-instantiating approach of
\cite{Montano-Rivas.McCasland.Dixon.ea:2012}, which forms part of a theory
exploration system), to others which are more closely related to theories of
human learning and discovery \cite{Piantadosi.Tenenbaum.Goodman:2012,
  mullerunderstanding}. Those based on finding patterns in data, such as
\cite{Wille:2005}, may be useful in tandem with our expression database, and
the results of value generators like \qcheck{}.

Since tools like \hspec{} already call out to third-party automated theorem
provers, and indeed our own ML4HS system uses the external Weka program, there
may also be merit in using external concept formation or conjecture generation
tools (or reimplementations of their ideas), in order to build up more structure
on top of that provided by the code we are exploring. For example, the
approaches taken by AM \cite{lenat1977automated, lenat1979automated}, Graffiti
\cite{delavina2005some, delavina2005graffiti} and HR
\cite{colton1999automatic, colton2000agent} could be used alongside those of
\qspec{}.

@@

  \caption{The Wundt curve, reproduced from \cite{berlyne1970novelty}. The axes ``hedonic value'' and ``arousal potential'' are described as covering \textquote{reward value\dots preference or pleasure}, and \textquote{all the stimulus properties that tend to raise arousal, including novelty and complexity}, respectively.}

@@

\emph{Artificial curiosity} (AC) describes active learning systems which are
rewarded based on how interesting the input or data they discover is
\cite{schmidhuber2006developmental}. Although framed in the context of
\emph{reinforcement learning}, this reliance on interest is clearly relevant to
our theory exploration setting.

As an unsupervised learning task, AC has no access to labels or meanings
associated with its input; the only features it can learn are the structure and
relationships inherent in the data, in a similar way to our recurrent clustering
algorithm. The unifying principle of AC methods is to force systems away from
inputs which are not amenable to learning; either because they are so familiar
that there is nothing left to learn, or so unfamiliar that they are
unintelligible. The resulting behaviour is characterised by the \emph{Wundt
  curve} (shown in Figure \ref{fig:wundt}) \footnote{In practice, many measures
  avoid negative values for simplicity, in which cases we replace all negative
  points on the curve with zero.}, which has been used in psychology to explain
human aesthetics and preferences \cite{berlyne1970novelty}. This same behaviour
may be applicable to the theorems produced by a theory exploration system.

@@

A framework encompassing many examples of the explicit approach is given in
\cite{oudeyer2007intrinsic} in the context of reinforcement learning; for
comparison, many similar measures are surveyed in a data mining context in
\cite{geng2006interestingness}. Many more reinforcement learning examples can
be found in \cite{Kaplan2006, Lipson2007, Luciw2011, Macedo2000,
  Ramik.Sabourin.Madani:2013, Roa.Kruijff.Jacobsson:2009, Schmidhuber:1991,
  oudeyer2004intelligent}; whilst more general descriptions are given in
\cite{Schaul.Sun.Wierstra.ea:2011, Scott1989, maher2008achieving}, which may be
more amenable to our theory exploration setting.

@@

Another example of explicit artificial curiosity is given in
\cite{Hester.Stone:2012}, where world states which cause \emph{disagreement}
among a population of decision trees (a \emph{random forest}
\cite{randomforests}) are considered interesting. Since the models make
stochastic predictions, the disagreement follows a Wundt curve as the complexity
of state transitions increases: for parts of the state space which have been
fully learned, the models will agree on accurate predictions; for parts which
are unlearnable, the models cannot infer any structure, and will converge to
reporting the same average value. Whilst the latter predictions may not be
\emph{accurate}, they will be \emph{in agreement}, hence pushing down the
interestingness of states which are too complex.

Many examples of the implicit case are based on \emph{coevolution}: rewarding
one part of the system for exploiting another part, and vice versa. In
\cite{Schmidhuber1999} a pair of learning algorithms place virtual ``bets'' on
the outcome of actions, and the winner is rewarded at the expense of the
loser. Due to the risk involved, each algorithm will only bet when it is
confident in its prediction, and bets will only be actioned when each algorithm
is confident in a \emph{different} outcome. The overall behaviour of this system
is therefore similar to the explicit measure of disagreement used in the random
forest example. In terms of theory exploration, such a scheme could be used to
find theorems which are \emph{non-obvious}, and hence informative in some way to
the user.

Another case is the ``darwinian brain'' of Fernando et
al. \cite{fernando2013design1, fernando2013design2}. This coevolves a
population of problem generators and problem solvers, rewarding the solvers
based on their speed, and rewarding the generators based on the \emph{variance}
of the solvers' speed. This avoids trivial problems (which all solvers can
quickly overcome) and complex problems (which no solver can manage), and focuses
on those with the most possibility for learning. It is easy to imagine such a
general architecture being populated by conjecture generators and theorem
provers to form a theory exploration system.

@@

Coevolution is a form of \emph{evolutionary computation}; an umbrella term for
heuristic search algorithms which mimic the process of evolution by natural
selection among a population of candidate solutions
\cite{back1997evolutionary}. Whilst \emph{genetic algorithms} are perhaps the
most well-known instance of evolutionary computation, their use of
\emph{strings} to represent solutions causes complications when comparing to a
domain like theory exploration, where recursive structures of unbounded depth
arise. Thankfully these problems are not insurmountable, for example
\emph{genetic programming} can operate on tree-structures natively
\cite{banzhaf1998genetic}, which makes evolutionary computation a useful source
of ideas for reuse in our theory exploration setting (there are also precedents
for using evolutionary computation in a theorem proving domain
\cite{spector2008genetic}).

@@

Pure exploration (i.e. for its own sake) has been studied in evolutionary
computation for two main reasons: \emph{artificial life} and \emph{deceptive
  problems}. The former attempts to gain insight into the nature of life and
biology through competition over limited resources. Whilst this may have utility
in resource allocation, e.g. efficient scheduling of a portfolio of ATP
programs, there is no direct connection to interestingness in theory
exploration, so we will not consider it further (note that similar
resource-usage ideas can also be found in the literature on \emph{artificial
  economies}, e.g. \cite{baum2000evolution}).

On the other hand, work on deceptive problems is highly relevant, as it has lead
to studying various notions of intrinsic fitness, which are analogous to the
interestingness measures we want. Deceptive problems are those where
\textquote{pursuing the objective may prevent the objective from being reached}
\cite{lehman2011abandoning}, which is caused by the fitness (objective)
function having many local optima which are easy to find (e.g. by hill
climbing), but few global optima which are hard to find. Many approaches try to
avoid deception by augmenting the given fitness function to promote
\emph{diversity} and \emph{novelty}, such as \emph{niching methods}
\cite{sareni1998fitness}.

@@

Traditional selection methods are biased towards high fitness individuals (this
is especially clear for truncation). Alternative schemes have been proposed
which favour diversity \emph{at the expense of} fitness. For example, the
fitness uniform selection scheme (FUSS) \cite{hutter2002fitness} selects a
target fitness $f_t$ uniformly from the interval
$\left[ f_{min}, f_{max} \right]$ between the highest and lowest of the
population. An individual $s$ is then selected with fitness closest to $f_t$,
i.e. $s = \argmin\limits_{x} \lvert f(x) - f_t \rvert$

@@

Once we begin this process of augmenting fitness functions, or abandoning their
use as objectives, an obvious question arises: what happens if our new function
contains nothing of the original? This kind of pure exploration scenario leads
to a variety of ideas for \emph{instrinsic} fitness, such as novelty
\cite{lehman2011abandoning}, which can lead to learning useful ``stepping
stones'' even in objective-driven domains. Such intrinsic notions of fitness are
direct analogues of the interestingness measures we seek for theory exploration.
